Modelling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak

### 2020-07-22 ExperimentalData/DIIID/145380/Paper/paper.tex
Latest version of paper

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding rational surfaces, which is the prediction of nonlinear theory.
In the second case, the natural frequencies are 
determined by the guiding-center flows at the rational surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  rational surfaces, which is the prediction of linear theory.
The second case gives  the best agreement between the simulations and the  experimental observations.   The third case gives by far the worst agreement between the simulations 
and the  observations. The first case only leads to partial agreement between  the  simulations and the  observations. However, this lack of complete agreement
may be a consequence of using an inaccurate guess for the neutral particle distribution in the pedestal.  
\end{abstract}

@@

According to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the rational surface (see the discussion in Sect.~\ref{natural}).\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\perp\,i} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_E({\mit\Psi}_p) &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $e$ is the magnitude of the electron charge, ${\mit\Psi}_p$  the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, and $n$ the toroidal mode number of the RMP.  Moreover, $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, 
equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Sect.~B of Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. 
The right-hand side of Eq.~(\ref{e2}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $n/m$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface. Moreover, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal magnetic flux (divided by $2\pi$).

@@

\section{EPEC Modeling of DIII-D Discharge \#145380}
\subsection{Plasma Equilibrium}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of DIII-D discharge \#145380. Note that the pedestal electron number density is relatively high ($n_{e\,{\rm ped}}\sim 3\times 10^{19}\,{\rm m}^{-3}$). 
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression are evident: the first extends from $t=2840$--$2980$ ms; the
second extends from $t=3320$--$3560$ ms; and the third extends from $t=3880$--$4200$ ms.

@@

Measured plasma equilibrium and
plasma profiles are available every $50$ ms. The equilibrium and profiles  at a given point in time are interpolated from
the measured equilibria and profiles. In fact, all quantities that depend on the plasma equilibrium and profiles are recalculated every 1 ms.  Data values from the
{\tt GPEC} code are available every 100 ms. As before, the {\tt GPEC} data values at a given point in time are interpolated from the
available {\tt GPEC} data values, and any quantity that depends on these values is recalculated every 1 ms. There is one major exception to
the aforementioned scheme. The theoretical model used in this paper actually requires plasma equilibria and profiles that are unaffected
by the applied RMP. However, there is clear evidence that the equilibria and the profiles (in particular) are significantly modified
within the three ELM-suppression windows (presumably because wide magnetic island chains are driven in the pedestal). [See Fig.~1(c).]
Hence, we do not use the equilibrium, profile, or {\tt GPEC}, data within the three ELM-suppression windows. Instead, we
interpolate across these windows using data taken on either side of the windows.

@@

Note that all quantities used in our theoretical model are derived from experimental measurements taken in DIII-D discharge \#145380, with the exception of the neutral particle data. 
Unfortunately, no neutral particle data is available for DIII-D discharge \#145380.
Hence, we employ neutral particle data derived from a much earlier discharge (see Sect.~III.D of Ref.~\onlinecite{rf2}).\cite{neutral}
In future, data from the new {\tt LLAMA} (Lyman Alpha Measurement Apparatus) system will be routinely available in DIII-D discharges.\cite{llama}
Combining this data  with modeling from the SOLPS-ITER code\,\cite{solps} ought to generate sufficient information to fully determine the
neutral particle data required by the {\tt EPEC} code.

@@

\section*{Acknowledgements}
The author would like to thank Q.M.~Hu and R.~Nazikian for providing the experimental data used in this paper. The
author would also like to thank J.-K.~Park and N.C.~Logan for providing guidance on how to run the {\tt GPEC} code.
This research was directly funded by the U.S.\ Department of Energy, Office of Science, Office of Fusion Energy Sciences,  under contract DE-FG02-04ER54742, and
incorporates work funded by the U.S.\ Department of Energy, Office of Science, Office of Fusion Energy Sciences, using the DIII-D National Fusion Facility, a DOE Office of Science user facility, under contract DE-FC02-04ER54698.

>>>>

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding rational surfaces, which is the prediction of nonlinear theory.
In the second case, the natural frequencies are 
determined by the guiding-center flows at the rational surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  rational surfaces, which is the prediction of linear theory.
The second case gives  the best agreement between the simulations and the  experimental observations.   The third case gives by far the worst agreement between the simulations 
and the  observations. The first case only leads to partial agreement between  the  simulations and the  observations. However, this lack of complete agreement
may be a consequence of using an inaccurate assumption for the neutral particle distribution in the pedestal.  
\end{abstract}

@@

According to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the rational surface (see the discussion in Sect.~\ref{natural}).\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\perp\,i} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_E({\mit\Psi}_p) &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $e$ is the magnitude of the electron charge, ${\mit\Psi}_p$  the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, and $n$ the toroidal mode number of the RMP.  Moreover, $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, 
equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Sect.~B of Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. 
The right-hand side of Eq.~(\ref{e2}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $m/n$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface. Moreover, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal magnetic flux (divided by $2\pi$).

@@

\section{EPEC Modeling of DIII-D Discharge \#145380}
\subsection{Plasma Equilibrium}
DIII-D discharge \#145380 is an ITER-Similar-Shape (ISS), ELMing, H-mode discharge, with a toroidal magnetic field $B_T=-1.9\,{\rm T}$, in which the
plasma current, $I_p$, is slowly ramped over a 2 second interval in order to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of DIII-D discharge \#145380. 
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression are evident: the first extends from $t=2840$--$2980$ ms; the
second extends from $t=3320$--$3560$ ms; and the third extends from $t=3880$--$4200$ ms.

@@

Measured plasma equilibrium and
plasma profiles are reconstructed every $50$ ms. The equilibrium and profiles  at a given point in time are interpolated from
the reconstructed equilibria and profiles. In fact, all quantities that depend on the plasma equilibrium and profiles are recalculated every 1 ms.  Data values from the
{\tt GPEC} code are calculated every 100 ms. As before, the {\tt GPEC} data values at a given point in time are interpolated from the
calculated {\tt GPEC} data values, and any quantity that depends on these values is recalculated every 1 ms. There is one major exception to
the aforementioned scheme. The theoretical model used in this paper actually requires plasma equilibria and profiles that are unaffected
by the applied RMP. However, there is clear evidence that the equilibria and the profiles (in particular) are significantly modified
within the three ELM-suppression windows (presumably because wide magnetic island chains are driven in the pedestal). [See Fig.~1(c).]
Hence, we do not use the equilibrium, profile, or {\tt GPEC}, data within the three ELM-suppression windows. Instead, we
interpolate across these windows using data taken on either side of the windows.

@@

Note that all quantities used in our theoretical model are derived from experimental measurements taken in DIII-D discharge \#145380, with the exception of the neutral particle data. 
Unfortunately, no neutral particle data is available for DIII-D discharge \#145380.
Hence, we employ neutral particle data derived from a much earlier discharge (see Sect.~III.D of Ref.~\onlinecite{rf2}).\cite{neutral}
In future, data from the new {\tt LLAMA} (Lyman Alpha Measurement Apparatus) system will be routinely available in DIII-D discharges.\cite{llama}
Combining this data  with modeling from a neutral particle simulation code such as, for example, the  SOLPS-ITER code\,\cite{solps} ought to generate sufficient information to fully determine the
neutral particle data required by the {\tt EPEC} code.

@@

\section*{Acknowledgements}
The author would like to thank Q.M.~Hu and R.~Nazikian for providing the experimental data used in this paper. The
author would also like to thank J.-K.~Park and N.C.~Logan for providing guidance on how to run the {\tt GPEC} code. Finally, the
author would like to thank C.~Paz-Solden and B.A.~Grierson for helpful comments during the preparation of this paper.

This research was directly funded by the U.S.\ Department of Energy, Office of Science, Office of Fusion Energy Sciences,  under contract DE-FG02-04ER54742, and
incorporates work funded by the U.S.\ Department of Energy, Office of Science, Office of Fusion Energy Sciences, using the DIII-D National Fusion Facility, a DOE Office of Science user facility, under contract DE-FC02-04ER54698.
